\textbf{Evaluation of machine learning methods for 16S rRNA gene data}

### 2019-07-05 submission/manuscript.tex
sent to Joshua:

Prediction of health outcomes using ML is rapidly being adopted by
human-associated microbiome studies. However, the developed ML so far
are overoptimistic in terms of predictive performance and validity.
Without rigorous ML pipelines, we cannot trust ML models. Before we can
speed up progress, we need to slow down, define and start using good ML
practices.

@@

Machine learning (ML) methods are effective at recognizing and
highlighting patterns in complex microbial datasets. Therefore,
researchers have started to explore the utility of ML models that use
microbiota associated biomarkers to predict human health and to
understand the microbial ecology of diseases such as liver cirrhosis,
colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2
diabetes and others (1--11). However, currently the field's use of ML
lacks clarity and consistency on which methods are used and how these
methods are implemented {[}(12); dadkhah\_gut\_2019{]}. More notably, we
commonly see ML practices such as using ML pipelines where there is no
seperate held-out test dataset to evaluate model performance or
reporting few or only the best outcomes of cross-validation. Even when
there are seperate testing sets to evaluate model performance, there are
large differences between cross-validation and testing performances that
indicate overfitting as well as large confidence intervals for testing
performances (4, 13--19). Moreover, there is a lack of discussion on why
a particular ML model is utilized. Recently, there is a trend towards
using more complex ML models such as random forest, extreme gradient
boosting and neural networks without a discussion on if and how much
model interpretibility is necessary for the study (11, 20--22). Black
box machine learning models require posthoc explanations to determine
the feature importances in making a prediction. These explanations can
be misleading and at times unreliable when making high-stake decisions
about someone's health (23). The lack of transparency on model selection
and interpretation as well as flawed modeling methods negatively impact
model validity and reproducibiliy. We need to strive toward better
machine learning practices by (1) implementing rigourous machine
learning pipelines and (2) selecting ML models that reflect the goal of
the study as it will inform our expectations of model accuracy,
complexity, interpretibility and computational efficiency.

To showcase a rigorous ML pipeline and to shed light on how much ML
model selection can affect modeling results, we performed an empirical
analysis comparing several different ML models using the same dataset
and the same ML pipeline. We used a previously published colorectal
cancer (CRC) study (3) which had fecal 16S rRNA gene sequences from 490
patients. CRC is a type of cancer which the human-associated human
microbiome is hypothesized to directly contribute to its development and
fecal 16S rRNA gene sequences have been used to detect CRC. We built ML
models using fecal 16S rRNA gene sequences to predict healthy patients
versus patients with colorectal lesions that were identified by
colonoscopy as screen relevant neoplasias (SRN). The study had 261
normal and 229 SRN samples. We established modeling pipelines for
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost. Our ML pipeline utilized held-out test data
to evaluate predictive performance and generalizability of each ML
model. We used the area under the receiver operating characteristic
curve (AUROC) as the predictive performance metric. The median test
AUROC varied from 0.601 to 0.695. Random forest had the highest median
AUROC for detecting SRN. Despite its simplicity, the L2-regularized
logistic regression was second best in predictive performance. In terms
of computational efficiency, L1 SVM with linear kernel trained the
fastest (0.202 hours, std ± 0.028), while XGBoost took the longest
(155.104 hours, std ± 0.959). We also found that depending on how the
data is split to create a held-out test set, the AUROC values of a ML
model can vary up to 0.295 which highlighted the importance of
performing many randomized modeling runs. This study established
standards for microbiome-associated ML models and underscored the
importance of model selection.

@@

We used a cohort of 490 patients with 261 cases of SRN. For each
patient, we had 6920 features (fecal bacterial abundances) and a
two-class label that defines their colonic health (having colorectal
lesions that are identified as SRN or normal). All the cases were
independently labeled through colonoscopies. We established modeling
pipelines for a binary prediction task with L2-regularized logistic
regression, L1 and L2 support vector machines (SVM) with linear and
radial basis function kernels, a decision tree, random forest and
extreme gradient boosted decision tree (XGBoost) to emphasize the
differences in model accuracy, complexity, interpretibility and
computational efficiency due to model selection.

@@

We established a ML pipeline where we train and validate each of the
seven models {[}Figure 1{]}. We randomly split the data into
training/validation and test sets so that the training/validation set
consisted of 80\% of the full dataset while the test set was composed of
the remaining data {[}Figure 1{]}. Since the cases are not uniformly
represented in the data, the initial data-split was stratified to
maintain the overall label distribution in both the training/validation
and test sets. Training/validation set consisted of 393 patients (209
SRN), while the test set was composed of 97 patients (52 SRN). The
training/validation data was used for training purposes and validation
of hyperparameter selection (i.e.~model selection), and the test set was
used for evaluation purposes. Validation of hyperparameter selection was
performed using repeated five-fold cross-validation on the
training/validation set {[}Figure 1{]}. Similar to the initial
data-split, five-fold cross-validation was also stratified to maintain
the overall label distribution on the training and validation sets. We
validated the cross-validation performances of each hyperparameter
setting over 100 randomizations and selected the best performing
hyper-parameter setting in terms of AUROC metric and trained the full
training/validation dataset {[}Figures S1 and S2{]}. We then used the
held-out test set to evaluate the prediction performance of each ML
model. The data-split, hyperparameter selection, training and testing
steps were repeated 100 times to get a reliable and robust reading of
model performance {[}Figure 1{]}.

@@

We evaluated the prediction performances of seven binary classification
models when applied to held-out test data using AUROC metric {[}Figure
1{]}. Random forest had significantly higher test AUROC values than the
other models for detecting SRNs when AUROC values were compared to the
other six by Wilcoxon rank sum test (p \textless{} 0.01). The median
AUROC of the random forest model was 0.695 (IQR 0.044). L2-regularized
logistic regression, XGBoost, L2-regularized SVM with linear and radial
basis function kernel AUROC values were not significantly different from
one another. They had median AUROC values of 0.68 (IQR 0.055), 0.679
(IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR 0.056) respectively. L1
SVM with linear kernel and decision tree had significantly lower AUROC
values than the other ML models with median AUROC of 0.65 (IQR 0.066)
and 0.601 (IQR 0.059), respectively {[}Figure 2{]}.

For each model, we compared the median cross-validation AUROC to the
median testing AUROC. The difference between the two should be low to
suggest the model is not overfitting despite the large number of
features. The largest difference between the two was in L1 SVM with
linear kernel, 0.021, followed by SVM with radial basis function kernel
and decision tree with a difference of 0.007 and 0.006, respectively
{[}Figure 2{]}.

@@

The ML models we built using L2-regularized logistic regression, L1 and
L2 support vector machines (SVM) with linear and radial basis function
kernels, a decision tree, random forest and XGBoost decrease in
interpretibility as they increase in complexity. We interpreted L1 and
L2 SVM with linear kernel and L2 logistic regression using the feature
weights of the trained models. We ranked the absolute weights of all the
OTUs for each data-split. We calculated the median ranks of these
features over the 100 data-splits. In the three linear models, OTUs that
had the largest median ranks and drove the detection of SRNs belong to
families \emph{Lachnospiraceae}, and \emph{Ruminococcaceae} (OTU01239,
OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822, OTU00609),
genera \emph{Gamella} (OTU00426) and genera \emph{Peptostreptococcus}
(OTU00367) {[}Figure 3{]}. Some of the OTUs with the highest ranks were
shared among the linear models. We explained the feature importances in
non-linear models using permutation importance on the held-out test data
where we randomly permuted non-correlated features individually and
groups of correlated features together (see methods) to calculate the
effect of permuted OTUs or group of OTUs on testing AUROC. The top 5
OTUs with the largest negative impact on testing AUROC overlapped in
tree-based models {[}Figure 4{]}. Specifically, permuting
\emph{Peptostreptococcus} (OTU00367) abundances randomly, dropped the
predictive performances the most in all tree-based methods {[}Figure
S3{]}. Decision tree, random forest and XGBoost models' predictive
performance dropped from 0.6 base testing AUROC median to 0.52, from
0.69 to 0.68 and from 0.68 to 0.65, respectively {[}Figure 4{]}.

@@

In this study we established a rigorous ML pipeline to use 16S rRNA
sequence counts to predict a binary health outcome. We set-up standards
for developing and evaluating ML models for microbiome data. We used a
held-out test set to illustrate the difference between cross-validation
and testing AUROC values. When the differences between cross-validation
and test performance is low, this suggets the models are not overfit and
that they will perform similar with similar data. In all seven models,
the differences in cross-validation and testing AUROC values did not
exceed 0.021 which suggests that these models will be able to test
similar new data. We performed the initial 80\%-20\% random datasplit
100 times in our ML pipeline. The repetition of the initial data-split
to create a held-out test set is a crucial step in the ML pipeline to
develop generalizable ML models and to report reliable performance
metrics for a ML model. Depending on the data-split, there is the chance
of being overoptimistic about the predictive performance of a model. In
our study, we showed that there was variability in AUROC values between
different random data-splits in each of the models we tested. Our
results showed that the testing AUROC values varied 0.23 on average,
depending on the data-split. We used AUROC metric instead of accuracy in
our study to evaluate predictive performance of the ML models. AUROC is
always random at the value 0.5 and is a robust metric when a dataset is
imbalanced. We performed a full grid search for hyperparameter settings
when building a ML model. It is not enough to use the default
hyperparameter settings when using previously developed ML packages in
programming languages like R, Python and Matlab. In the example of
L1-regularized SVM with linear kernel (Figure S1), the model showed
large variability between different regularization coefficients (C) and
was susceptible to performing poorly if the wrong regularization
coefficient was assigned to the model by default.

Our results showed that we should choose to use ML models based on the
goal of the study and our expectations of predictive performance,
interpretibility and computational burden. In terms of predictive
performance, random forest model had testing AUROC values statistically
significantly larger than others. However the second best model was
L2-regularized logistic regression following random forest with a median
AUROC difference of only 0.015. In terms of interpretation, random
forest was a more complex ML model and it could only be explained using
methods such as permutation importance. On the other hand,
L2-regularized logistic regression was easy to interpret
(i.e.~regression coefficients of the trained model). If the goal of a
study is to learn the ecology behind a disease and to identify microbial
biomarkers of a disease, then easily interpretable models would be
better suited. It is also important to consider the computational burden
of developing ML models. Random forest model was slow to train 83.2
hours whereas L2-regularized logistic regression trained in 12 minutes.
The dataset we used for our study had 490 samples. Another criteria when
choosing a ML model is the generalizability of it. The generalization
performance of ML models depends on size. The more complex the model,
the more data it will need. Microbiome studies that have small sample
sizes should use less complex models such as L2-regularized logistic
regression.

@@

\textbf{Data collection and study population.} The data used for this
analysis are stool bacterial abundances and clinical information of the
patients recruited by Great Lakes-New England Early Detection Research
Network study. These data were obtained from Sze et al (24). The stool
samples were provided by recruited adult participants who were
undergoing scheduled screening or surveillance colonoscopy.
Colonoscopies were performed and fecal samples were collected from
participants in four locations: Toronto (ON, Canada), Boston (MA, USA),
Houston (TX, USA), and Ann Arbor (MI, USA). Patients' colonic health was
labeled by colonoscopy with adequate preparation and tissue
histopathology of all resected lesions. Patients with an adenoma greater
than 1 cm, more than three adenomas of any size, or an adenoma with
villous histology were classified as advanced adenoma. Study had 172
patients with normal colonoscopies, 198 with adenomas and 120 with
carcinomas. Of the 198 adenomas, 109 were identified as advanced
adenomas. Stool provided by the patients was used for 16S rRNA gene
sequencing to measure bacterial population abundances. The bacterial
abundance data was generated by Sze et al, by processing 16S rRNA
sequences in Mothur (v1.39.3) using the default quality filtering
methods, identifying and removing chimeric sequences using VSEARCH and
assigning to OTUs at 97\% similarity using the OptiClust algorithm
(25--27).

@@

The colonic health of the patient was defined as two encompassing
classes; Normal or Screen Relevant Neoplasias (SRNs). Normal class
includes patients with non-advanced adenomas or normal colons whereas
SRN class includes patients with advanced adenomas or carcinomas. The
bacterial abundances are the features used to predict colonic health of
the patients. Bacterial abundances are discrete data in the form of
Operational Taxonomic Unit (OTU) counts. OTU counts were set to the size
of our smallest sample and were subsampled at the same distances. They
were then transformd by scaling to a {[}0-1{]} range.

@@

For L2-regularized logistic regression, L1 and L2 support vector
machines (SVM) with linear and radial basis function kernels we tuned
the \textbf{cost} hyperparameter which determines the regularization
strength where smaller values specify stronger regularization. For SVM
with radial basis function kernel we also tuned \textbf{sigma}
hyperparameter which determines the reach of a single training instance
where for a high value of sigma, the SVM decision boundary will be
dependent on the points that are closest to the decision boundary. For
the decision tree model, we tuned the \textbf{depth of the tree} where
deeper the tree, the more splits it has. For random forest, we tuned the
\textbf{number of features} to consider when looking for the best tree
split. For XGBoost, we tuned for \textbf{learning rate} and the
\textbf{fraction of samples} to be used for fitting the individual base
learners.For hyperparameter selection, we started with a granular grid
search. Then we narrowed and fine-tuned the range of each
hyperparameter. The range of the grid depends on the ML task and ML
model. A full grid search needs to be performed to avoid variability in
testing performance. We can use hyper-band to help us with our
hyperparameter selection (28).

The computational burden during model training due to model complexity
was reduced by parallelizing segments of the ML pipeline. In this study
we have parallelized each data-split which allowed 100 data-splits to be
processed through the ML pipeline at the same time for each model. We
can further parallelize the cross-validation step for each
hyper-parameter setting.

@@

\includegraphics{Figure_1} \textbf{Figure 1. Machine learning pipeline
showing predictive model training and evaluation flowchart. } We split
the data 80\%/20\% stratified to maintain the overall label
distribution, performed five-fold cross-validation on the training data
to select the best hyperparameter setting and then using these
hyperparameters to train all of the training data. The model was
evaluated on a held-out set of data (not used in selecting the model).
Abbreviations: AUROC, area under the receiver operating characteristic
curve \newpage
\includegraphics{Figure_2.png} \textbf{Figure 2. Generalization and
classification performance of ML models using AUROC values of all cross
validation and testing performances. } The median AUROC for diagnosing
individuals with SRN using bacterial abundances was higher than chance
(depicted by horizontal line at 0.50) for all the ML models.
Discriminative performance of random forest model was higher than other
ML models. The boxplot shows quartiles at the box ends and the
statistical median as the horizontal line in the box. The whiskers show
the farthest points that are not outliers. Outliers are data points that
are not within 3/2 times the interquartile ranges. Abbreviations: SRN,
screen-relevant neoplasias; AUROC, area under the receiver operating
characteristic curve; SVM, support vector machine; XGBoost, extreme
gradient boosting \newpage
\includegraphics{Figure_3.png} \textbf{Figure 3. Interpretation of the
linear ML models.} (A) L2 logistic regression coefficients (B) L1 SVM
with linear kernel feature weights (C) L2 SVM with linear kernel feature
ranks. The feature ranks of the most important five OTUs in each
data-split are shown here. Similar OTUs had the largest impact on the
predictive performance of L2 logistic regression and L2 SVM with linear
kernel. Abbreviations: SVM, support vector machine; OTU, Operational
Taxonomic Unit. \newpage
\includegraphics{Figure_4.png} \textbf{Figure 4. Explanation of the
non-linear ML models.} (A) SVM with radial basis kernel (B) decision
tree (C) random forest (D) XGboost feaure importances were explanied
using permutation importance using held-out test set. The gray rectangle
and the dashed line show the IQR range and median of the base testing
AUROC without any permutation performed. For all the tree-based models,
a \emph{Peptostreptococcus} species (OTU00367) had the largest impact on
predictive performance of the model. Abbreviations: SVM, support vector
machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU,
Operational Taxonomic Unit. \newpage
\includegraphics{Figure_5.png} \textbf{Figure 5. Computational
efficiency of seven ML models.} The training times for training and
testing of each data-split showed the differences in computational
efficieny of the seven models. The median traintime in hours was the
highest for XGBoost and shortest for L2 logistic regression. The boxplot
shows quartiles at the box ends and the statistical median as the
horizontal line in the box. The whiskers show the farthest points that
are not outliers. Outliers are data points that are not within 3/2 times
the interquartile ranges. Abbreviations: AUROC, area under the receiver
operating characteristic curve; SVM, support vector machine; XGBoost,
extreme gradient boosting.\\
\newpage
\includegraphics{Figure_S1.png} \textbf{Figure S1. Hyperparameter
setting performances for linear models.} Abbreviations: \newpage
\includegraphics{Figure_S2.png} \textbf{Figure S2. Hyperparameter
setting performances for non-linear models.} Abbreviations: \newpage

@@

\hypertarget{ref-flemer_oral_2018}{}
13. \textbf{Flemer B}, \textbf{Warren RD}, \textbf{Barrett MP},
\textbf{Cisek K}, \textbf{Das A}, \textbf{Jeffery IB}, \textbf{Hurley
E}, \textbf{O`Riordain M}, \textbf{Shanahan F}, \textbf{O`Toole PW}.
2018. The oral microbiota in colorectal cancer is distinctive and
predictive. Gut \textbf{67}:1454--1463.
doi:\href{https://doi.org/10.1136/gutjnl-2017-314814}{10.1136/gutjnl-2017-314814}.

>>>>

Prediction of health outcomes using ML is rapidly being adopted by
human-associated microbiome studies. However, the developed ML models so
far are overoptimistic in terms of validity and predictive performance.
Without rigorous ML pipelines, we cannot trust ML models. Before we can
speed up progress, we need to slow down, define and start using good ML
practices. \newpage

@@

Machine learning (ML) methods are effective at recognizing and
highlighting patterns in complex microbial datasets. Therefore,
researchers have started to explore the utility of ML models that use
microbiota associated biomarkers to predict human health and to
understand the microbial ecology of diseases such as liver cirrhosis,
colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2
diabetes and others (1--11). However, currently the field's use of ML
lacks clarity and consistency on which methods are used and how these
methods are implemented (12, 13). More notably, we commonly see flawed
ML practices such as using ML pipelines where there is no seperate
held-out test dataset to evaluate model performance or reporting few or
only the best outcomes of cross-validation. Even when there are seperate
testing sets to evaluate model performance, there are large differences
between cross-validation and testing performances that indicate
overfitting as well as large confidence intervals for testing
performances (4, 14--20). Moreover, there is a lack of discussion on why
a particular ML model is utilized. Recently, there is a trend towards
using more complex ML models such as random forest, extreme gradient
boosting and neural networks without a discussion on if and how much
model interpretibility is necessary for the study (11, 21--23). Black
box machine learning models require posthoc explanations to determine
the feature importances in making a prediction. These explanations can
be misleading and at times unreliable when making high-stake decisions
about someone's health (24). The lack of transparency on model selection
and interpretation as well as flawed modeling methods negatively impact
model validity and reproducibiliy. We need to strive toward better
machine learning practices by (1) implementing rigourous machine
learning pipelines and (2) selecting ML models that reflect the goal of
the study as it will inform our expectations of model accuracy,
complexity, interpretibility and computational efficiency.

To showcase a rigorous ML pipeline and to shed light on how much ML
model selection can affect modeling results, we performed an empirical
analysis comparing several different ML models using the same dataset
and the same ML pipeline. We used a previously published colorectal
cancer (CRC) study (3) which had fecal 16S rRNA gene sequences from 490
patients. CRC is a type of cancer which the human-associated human
microbiome is hypothesized to directly contribute to its development and
fecal 16S rRNA gene sequences have been used to detect CRC. We built
seven ML models using fecal 16S rRNA gene sequences to predict healthy
patients versus patients with colorectal lesions that were identified by
colonoscopy as screen relevant neoplasias (SRN). The study had 261
normal and 229 SRN samples. We established modeling pipelines for
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost. Our ML pipeline utilized held-out test data
to evaluate predictive performance and generalizability of each ML
model. We used the area under the receiver operating characteristic
curve (AUROC) as the predictive performance metric. The median test
AUROC varied from 0.601 to 0.695. Random forest had the highest median
AUROC for detecting SRN. Despite its simplicity, the L2-regularized
logistic regression was second best in predictive performance. In terms
of computational efficiency, L1 SVM with linear kernel trained the
fastest (0.202 hours, std ± 0.028), while XGBoost took the longest
(155.104 hours, std ± 0.959). We also found that depending on how the
data is split to create a held-out test set, the AUROC values of a ML
model can vary up to 0.295 which highlighted the importance of
performing many randomized modeling runs. This study established
standards for microbiome-associated ML models and underscored the
importance of model selection.

@@

We used a cohort of 490 patients with 261 cases of SRN. For each
patient, we had 6920 features (fecal bacterial abundances) and a
two-class label that defines their colorectal health (having colorectal
lesions that are identified as SRN or normal). All the cases were
independently labeled through colonoscopies. We established modeling
pipelines for a binary prediction task with L2-regularized logistic
regression, L1 and L2 support vector machines (SVM) with linear and
radial basis function kernels, a decision tree, random forest and
extreme gradient boosted decision tree (XGBoost) to emphasize the
differences in model accuracy, complexity, interpretibility and
computational efficiency due to model selection.

@@

We established a ML pipeline where we train and validate each of the
seven models {[}Figure 1{]}. We randomly split the data into
training/validation and test sets so that the training/validation set
consisted of 80\% of the full dataset while the test set was composed of
the remaining data {[}Figure 1{]}. Since the cases are not uniformly
represented in the data, the initial data-split was stratified to
maintain the overall label distribution in both the training/validation
and test sets. Training/validation set consisted of 393 patients (209
SRN), while the test set was composed of 97 patients (52 SRN). The
training/validation data was used for training purposes and validation
of hyperparameter selection (i.e.~model selection), and the test set was
used for evaluation purposes. Validation of hyperparameter selection was
performed using repeated five-fold cross-validation on the
training/validation set {[}Figure 1{]}. Similar to the initial
data-split, five-fold cross-validation was also stratified to maintain
the overall label distribution on the training and validation sets. We
validated the cross-validation performances of each hyperparameter
setting over 100 randomizations and selected the best performing
hyperparameter setting in terms of AUROC metric and trained the full
training/validation dataset {[}Figures S1 and S2{]}. We then used the
held-out test set to evaluate the prediction performance of each ML
model. The data-split, hyperparameter selection, training and testing
steps were repeated 100 times to get a reliable and robust reading of
model performance {[}Figure 1{]}.

@@

We evaluated the prediction performances of seven binary classification
models when applied to held-out test data using AUROC metric {[}Figure
2{]}. Random forest had significantly higher test AUROC values than the
other models for detecting SRNs when AUROC values were compared to the
other six by Wilcoxon rank sum test (p \textless{} 0.01). The median
AUROC of the random forest model was 0.695 (IQR 0.044). L2-regularized
logistic regression, XGBoost, L2-regularized SVM with linear and radial
basis function kernel AUROC values were not significantly different from
one another. They had median AUROC values of 0.68 (IQR 0.055), 0.679
(IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR 0.056) respectively. L1
SVM with linear kernel and decision tree had significantly lower AUROC
values than the other ML models with median AUROC of 0.65 (IQR 0.066)
and 0.601 (IQR 0.059), respectively {[}Figure 2{]}.

For each model, we compared the median cross-validation AUROC to the
median testing AUROC. The difference between the two should be low to
suggest the model is not overfitting despite the large number of
features. The largest difference between the two was 0.021 in L1 SVM
with linear kernel, followed by SVM with radial basis function kernel
and decision tree with a difference of 0.007 and 0.006, respectively
{[}Figure 2{]}.

@@

The ML models we built using L2-regularized logistic regression, L1 and
L2 support vector machines (SVM) with linear and radial basis function
kernels, a decision tree, random forest and XGBoost decrease in
interpretibility as they increase in complexity. We interpreted L1 and
L2 SVM with linear kernel and L2 logistic regression using the feature
weights of the trained models. We ranked the absolute weights of all the
OTUs for each data-split {[}Figure 3{]}. We calculated the median ranks
of these features over the 100 data-splits. In the three linear models,
OTUs that had the largest median ranks and drove the detection of SRNs
belonged to families \emph{Lachnospiraceae}, and \emph{Ruminococcaceae}
(OTU01239, OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822,
OTU00609), genera \emph{Gamella} (OTU00426) and genera
\emph{Peptostreptococcus} (OTU00367) {[}Figure 3{]}. Some of the OTUs
with the highest ranks were shared among the linear models. We explained
the feature importances in non-linear models using permutation
importance on the held-out test data where we randomly permuted
non-correlated features individually and groups of correlated features
together (see methods) to calculate the effect of permuted OTUs or group
of OTUs on testing AUROC. The top 5 OTUs with the largest negative
impact on testing AUROC overlapped in tree-based models {[}Figure 4{]}.
Specifically, permuting \emph{Peptostreptococcus} (OTU00367) abundances
randomly, dropped the predictive performances the most in all tree-based
methods {[}Figure S3{]}. Decision tree, random forest and XGBoost
models' predictive performance dropped from 0.6 base testing AUROC
median to 0.52, from 0.69 to 0.68 and from 0.68 to 0.65, respectively
{[}Figure 4{]}.

@@

In this study we established a rigorous ML pipeline to use 16S rRNA
sequence counts to predict a binary health outcome. We set-up standards
for developing and evaluating ML models for microbiome data. First, we
used a held-out test set to illustrate the difference between
cross-validation and testing AUROC values. When the differences between
cross-validation and test performance is low, this suggets the models
are not overfit and that they will perform similar with similar data. In
all seven models, the differences in cross-validation and testing AUROC
values did not exceed 0.021 which suggests that these models will be
able to test similar new data. Second, we performed the initial
80\%-20\% random datasplit 100 times in our ML pipeline. The
randomization of the initial data-split to create a held-out test set is
a crucial step in the ML pipeline to develop generalizable ML models and
to report reliable performance metrics for a ML model. Depending on the
data-split, there is the chance of being overoptimistic about the
predictive performance of a model. In our study, we showed that there
was variability in AUROC values between different random data-splits in
each of the models we tested. Our results showed that the testing AUROC
values varied 0.23 on average, depending on the data-split.Third, we
used AUROC metric instead of accuracy in our study to evaluate
predictive performance of the ML models. AUROC is always random at the
value 0.5 and is a robust metric when a dataset is imbalanced. We also
performed a full grid search for hyperparameter settings when building a
ML model. It is not enough to use the default hyperparameter settings
when using previously developed ML packages in programming languages
such as R, Python and Matlab. In the example of L1-regularized SVM with
linear kernel (Figure S1), the model showed large variability between
different regularization coefficients (C) and was susceptible to
performing poorly if the wrong regularization coefficient was assigned
to the model by default.

Our results showed that we should choose to use ML models based on the
goal of the study and our expectations of predictive performance,
interpretibility and computational burden. In terms of predictive
performance, random forest model had testing AUROC values statistically
significantly larger than others. However the second best model was
L2-regularized logistic regression following random forest with a median
AUROC difference of only 0.015. In terms of interpretation, random
forest was a more complex ML model and it could only be explained using
methods such as permutation importance. On the other hand,
L2-regularized logistic regression was easy to interpret
(i.e.~regression coefficients of the trained model). If the goal of a
study is to learn the ecology behind a disease and to identify microbial
biomarkers of a disease, then easily interpretable models would be
better suited. It is also important to consider the computational burden
of developing ML models. Random forest model was slow to train 83.2
hours whereas L2-regularized logistic regression trained in 12 minutes.
Another criteria when choosing a ML model is generalizability. The
generalization performance of ML models depends on size. The more
complex the model, the more data it will need. The dataset we used for
our study had 490 samples, however microbiome studies that have smaller
sample sizes would benefit from using less complex models such as
L2-regularized logistic regression.

@@

\textbf{Data collection and study population.} The data used for this
analysis are stool bacterial abundances and clinical information of the
patients recruited by Great Lakes-New England Early Detection Research
Network study. These data were obtained from Sze et al (25). The stool
samples were provided by recruited adult participants who were
undergoing scheduled screening or surveillance colonoscopy.
Colonoscopies were performed and fecal samples were collected from
participants in four locations: Toronto (ON, Canada), Boston (MA, USA),
Houston (TX, USA), and Ann Arbor (MI, USA). Patients' colonic health was
labeled by colonoscopy with adequate preparation and tissue
histopathology of all resected lesions. Patients with an adenoma greater
than 1 cm, more than three adenomas of any size, or an adenoma with
villous histology were classified as advanced adenoma. Study had 172
patients with normal colonoscopies, 198 with adenomas and 120 with
carcinomas. Of the 198 adenomas, 109 were identified as advanced
adenomas. Stool provided by the patients was used for 16S rRNA gene
sequencing to measure bacterial population abundances. The bacterial
abundance data was generated by Sze et al, by processing 16S rRNA
sequences in Mothur (v1.39.3) using the default quality filtering
methods, identifying and removing chimeric sequences using VSEARCH and
assigning to OTUs at 97\% similarity using the OptiClust algorithm
(26--28).

@@

The colorectal health of the patient was defined as two encompassing
classes; Normal or Screen Relevant Neoplasias (SRNs). Normal class
includes patients with non-advanced adenomas or normal colons whereas
SRN class includes patients with advanced adenomas or carcinomas. The
bacterial abundances are the features used to predict colonic health of
the patients. Bacterial abundances are discrete data in the form of
Operational Taxonomic Unit (OTU) counts. OTU counts were set to the size
of our smallest sample and were subsampled at the same distances. They
were then transformd by scaling to a {[}0-1{]} range.

@@

For L2-regularized logistic regression, L1 and L2 support vector
machines (SVM) with linear and radial basis function kernels we tuned
the \textbf{cost} hyperparameter which determines the regularization
strength where smaller values specify stronger regularization. For SVM
with radial basis function kernel we also tuned \textbf{sigma}
hyperparameter which determines the reach of a single training instance
where for a high value of sigma, the SVM decision boundary will be
dependent on the points that are closest to the decision boundary. For
the decision tree model, we tuned the \textbf{depth of the tree} where
deeper the tree, the more splits it has. For random forest, we tuned the
\textbf{number of features} to consider when looking for the best tree
split. For XGBoost, we tuned for \textbf{learning rate} and the
\textbf{fraction of samples} to be used for fitting the individual base
learners.For hyperparameter selection, we started with a granular grid
search. Then we narrowed and fine-tuned the range of each
hyperparameter. The range of the grid depends on the ML task and ML
model. A full grid search needs to be performed to avoid variability in
testing performance. We can use hyper-band to help us with our
hyperparameter selection (29).

The computational burden during model training due to model complexity
was reduced by parallelizing segments of the ML pipeline. In this study
we have parallelized each data-split which allowed 100 data-splits to be
processed through the ML pipeline at the same time for each model. We
can further parallelize the cross-validation step for each
hyperparameter setting.

@@

\includegraphics{Figure_1} \textbf{Figure 1. Machine learning pipeline
showing predictive model training and evaluation flowchart. } We split
the data 80\%/20\% stratified to maintain the overall label
distribution, performed five-fold cross-validation on the training data
to select the best hyperparameter setting and then using these
hyperparameters to train all of the training data. The model was
evaluated on a held-out set of data (not used in selecting the model).
Abbreviations: AUROC, area under the receiver operating characteristic
curve \newpage
\includegraphics{Figure_2.png} \textbf{Figure 2. Generalization and
classification performance of ML models using AUROC values of all cross
validation and testing performances. } The median AUROC for diagnosing
individuals with SRN using bacterial abundances was higher than chance
(depicted by horizontal line at 0.50) for all the ML models.
Discriminative performance of random forest model was higher than other
ML models. The boxplot shows quartiles at the box ends and the
statistical median as the horizontal line in the box. The whiskers show
the farthest points that are not outliers. Outliers are data points that
are not within 3/2 times the interquartile ranges. Abbreviations: SRN,
screen-relevant neoplasias; AUROC, area under the receiver operating
characteristic curve; SVM, support vector machine; XGBoost, extreme
gradient boosting \newpage
\includegraphics{Figure_3.png} \textbf{Figure 3. Interpretation of the
linear ML models.} (A) L2 logistic regression coefficients (B) L1 SVM
with linear kernel feature weights (C) L2 SVM with linear kernel feature
ranks. The feature ranks of the most important five OTUs in each
data-split are shown here. Similar OTUs had the largest impact on the
predictive performance of L2 logistic regression and L2 SVM with linear
kernel. Abbreviations: SVM, support vector machine; OTU, Operational
Taxonomic Unit. \newpage
\includegraphics{Figure_4.png} \textbf{Figure 4. Explanation of the
non-linear ML models.} (A) SVM with radial basis kernel (B) decision
tree (C) random forest (D) XGboost feaure importances were explanied
using permutation importance using held-out test set. The gray rectangle
and the dashed line show the IQR range and median of the base testing
AUROC without any permutation performed. For all the tree-based models,
a \emph{Peptostreptococcus} species (OTU00367) had the largest impact on
predictive performance of the model. Abbreviations: SVM, support vector
machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU,
Operational Taxonomic Unit. \newpage
\includegraphics{Figure_5.png} \textbf{Figure 5. Computational
efficiency of seven ML models.} The training times for training and
testing of each data-split showed the differences in computational
efficieny of the seven models. The median traintime in hours was the
highest for XGBoost and shortest for L2 logistic regression. The boxplot
shows quartiles at the box ends and the statistical median as the
horizontal line in the box. The whiskers show the farthest points that
are not outliers. Outliers are data points that are not within 3/2 times
the interquartile ranges. Abbreviations: AUROC, area under the receiver
operating characteristic curve; SVM, support vector machine; XGBoost,
extreme gradient boosting.\\
\newpage
\includegraphics{Figure_S1.png} \textbf{Figure S1. Hyperparameter
setting performances for linear models.} (A) L2 logistic regression (B)
L1 SVM with linear kernel (C) L2 SVM with linear kernel mean
cross-validation AUROC values when different hyperparameters are used in
training the model. The differences in AUROC values when hyperparameters
change show that hyperparameter tuning is a crucial step in building a
ML model.

\includegraphics{Figure_S2.png} \textbf{Figure S2. Hyperparameter
setting performances for non-linear models.} (A) Decision tree (B)
Random forest (C) SVM with radial basis kernel (D) XGoost mean
cross-validation AUROC values when different hyperparameters are used in
training the model. The differences in AUROC values when hyperparameters
change show that hyperparameter tuning is a crucial step in building a
ML model. \newpage

\includegraphics{TableS1.pdf} \textbf{Table S1. ML model complexity and
hyperparameter range.} \newpage

@@

\hypertarget{ref-dadkhah_gut_2019}{}
13. \textbf{Dadkhah E}, \textbf{Sikaroodi M}, \textbf{Korman L},
\textbf{Hardi R}, \textbf{Baybick J}, \textbf{Hanzel D}, \textbf{Kuehn
G}, \textbf{Kuehn T}, \textbf{Gillevet PM}. 2019. Gut microbiome
identifies risk for colorectal polyps. BMJ Open Gastroenterology
\textbf{6}:e000297.
doi:\href{https://doi.org/10.1136/bmjgast-2019-000297}{10.1136/bmjgast-2019-000297}.
